# Non-Communicable Disease Burden (NCD)

```{r echo=FALSE, error=FALSE, message=FALSE}
# Load necessary packages
pacman::p_load(tidyverse, knitr, here, janitor)

# Import the non-communicable diseases data
data_ncd <- read_csv(here("data/burden-of-disease-ncd.csv"))
```

```{r echo=FALSE}

data_ncd_clean <- data_ncd %>% 
  rename(dalys = `DALYs (Disability-Adjusted Life Years) - Non-communicable diseases - Sex: Both - Age: Age-standardized (Rate)`) %>% 
  clean_names() %>% 
  filter(code == "DOM" | code == "PER" | code == "MEX") #another option is using %in% as in filter(code %in% c("DOM", "PER", "MEX"))

```


## Table of Estimates for NCD Burden Over Time

The next table shows the DALYs attributed to NCDs for the three countries from 1990 to 2019.

```{r echo=FALSE}
# Here render a table for the DALY burden over time for the three countries. 
# You should pivot the data to show each country in a separate column.
# Use kable() from the knitr package to render the table.

data_ncd_wide <- data_ncd_clean %>%
  select(-entity) %>% 
  pivot_wider(names_from = code,
              values_from = dalys)

ncd_table <- kable(data_ncd_wide)
ncd_table

```

## Plot Showing Trends in NCD Burden Over Time 

The plot shows the trends in DALYs attributed to NCDs per country.

```{r, echo=FALSE}
# Here, plot a line graph showing the trend of DALY burden for the chosen countries over time. You can use ggplot2 to create this plot. Each country's line should be a different color.

plot_ncd <- ggplot(data = data_ncd_clean,
                   mapping = aes(x = year,
                                 y = dalys,
                                 color = entity)) +
  geom_line() +
  labs(title = "Burden of NCDs over time",
       color = "Country",
       x = "Year",
       y = "DAlYs")

plot_ncd
```

## Summary of NCD Burden Findings

3 chosen from the Latin America and Caribean Region were chosen: Dominican Republic, Mexico and Peru.

The trend of NCD burden in Peru seems to be declining, while the one for Dominican Republic increases over time. 

In the case of Mexico, there has been a slow decline that has become slower in recent years.


